Open Clusters Characterization in Gaia DR2 Using ML Algorithms

### 2020-11-29 thesis/thesis.tex
thesis: Minor corrections

\usepackage{apacite}
\usepackage{caption}
\usepackage{eso-pic}
\usepackage{float}
\usepackage{graphicx}
\usepackage{minted}
\usepackage[nottoc]{tocbibind}
\usepackage{picture}
\usepackage{subcaption}

@@

Stellar open clusters (OC) \cite{janes1982open} are groups of stars gravitationally bound originated from a single molecular gas cloud.
Thus they share the same chemical composition and age, and they have similar relative positions and proper motion.
These astronomical objects are of fundamental importance to understand the spiral structure,
the dynamics and the chemical evolution of our galaxy.

@@

Young OC allow us to research stars formation regions and improve our understanding about the mechanisms that create those stars.
On the other side, older OC give us information about stellar processes and how the galactic disk evolves.
Some highly disturbed orbits could also provide evidence of recent merge events and accretion traces from outside the galaxy
\cite{cantat2016abundances}.

The study of OC has been pushed forward thanks to the huge and precise dataset from the Gaia mission
\cite{collaboration2016description} Gaia DR2 \cite{gaia2018gaia}, available since 2018.
This dataset has helped to review already known open clusters and to find new ones.

Stars that belong to a same OC share relative positions, so their parallax value is similar among them. They also share similar values of
proper motion, both in right ascension and declination. Another property, since these stars were born from the same gas cloud and at the same
temporary stage, their metallicity must be uniform. However, to take into account this last property, we face the inconvenient that this
parameter is not directly available in Gaia DR2 database. We will avoid this issue by looking at $E(b-r) / G_{mag}$ diagrams with well
defined isochrone curves \cite{bressan2012parsec}. These isochrone curves are derived from theoretical models that mainly involve metallicity
and mass/brightness from stars. With these models we can define important properties for the development of stars and their evolution in the
Hertzsprung–Russell (H-R) diagram. Examples of these curves are shown in Figure \ref{fig:examples_of_isochrones}.

@@

We will take advantage of these properties to validate that the open clusters obtained are right.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.9\textwidth]{../figures/theoretical_isochrones_in_hr_diagrams.pdf}
  \caption{Examples of theoretical isochrones in the HR diagram taken from \protect\citeA[p.~16]{bressan2012parsec}}
  \label{fig:examples_of_isochrones}
\end{figure}

@@

% TODO: Hay que desarrollar más los objetivos

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/pm_melotte_22.pdf}
      \caption{Proper motions}
      \label{fig:pm_melotte_22}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/parallax_melotte_22.pdf}
      \caption{Parallax}
      \label{fig:pm_vec_melotte_22}
    \end{subfigure}
  \end{subfigure}
  \caption{Open Cluster Melotte 22 (M45)}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/pm_ngc_6494.pdf}
      \caption{Proper motions}
      \label{fig:pm_ngc_6494}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/parallax_ngc_6494.pdf}
      \caption{Parallax}
      \label{fig:parallax_ngc_6494}
    \end{subfigure}
  \end{subfigure}
  \caption{Open Cluster NGC 6494}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.5\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/raw_pm_melotte_22.pdf}
  \end{subfigure}
  \caption{Melotte 22 proper motions}
  \label{fig:raw_pm_melotte_22}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/raw_parallax_melotte_22.pdf}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/raw_hr_diagram_melotte_22.pdf}
    \end{subfigure}
  \end{subfigure}
  \caption{Melotte 22 parallax and H-R diagram}
  \label{fig:raw_parallax_hr_diagram_mellote_22}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/features_melotte_22.pdf}
  \end{subfigure}
  \caption{Feature pairplot}
  \label{fig:features_melotte_22}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans_n2_pm_melotte_22.pdf}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans_n5_pm_melotte_22.pdf}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans_n8_pm_melotte_22.pdf}
    \end{subfigure}
  \end{subfigure}
  \medskip
  \begin{subfigure}{0.9\textwidth}
    \centering
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans_n2_parallax_melotte_22.pdf}
      \caption{N clusters = 2}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans_n5_parallax_melotte_22.pdf}
      \caption{N clusters = 5}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans_n8_parallax_melotte_22.pdf}
      \caption{N clusters = 8}
    \end{subfigure}
  \end{subfigure}
  \caption{K-Means comparisons with Melotte 22}
  \label{fig:kmeans_comparisons_melotte_22}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \begin{subfigure}[t]{.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans_pm_melotte_22.pdf}
      \caption{Proper Motion}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans_parallax_melotte_22.pdf}
      \caption{Parallax}
    \end{subfigure}
  \end{subfigure}
  \caption{K-Means model applied to Melotte 22}
  \label{fig:kmeans_melotte_22}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.5\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/kmeans_isochrone_melotte_22.pdf}
  \end{subfigure}
  \caption{Melotte 22 H-R diagram with K-Means characterization}
  \label{fig:kmeans_hr_diagram_melotte_22}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/dec_pm_melotte_22.pdf}
      \caption{Proper Motion}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/dec_parallax_melotte_22.pdf}
      \caption{Parallax}
    \end{subfigure}
  \end{subfigure}
  \caption{DEC model applied to Melotte 22}
  \label{fig:dec_melotte_22}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.5\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/dec_isochrone_melotte_22.pdf}
  \end{subfigure}
  \caption{Melotte 22 H-R diagram with DEC characterization}
  \label{fig:dec_hr_diagram_melotte_22}
\end{figure}

@@

% TODO: Las tecnologías empleadas para el desarrollo de las herramientas,
% la organización del piloto, cómo ha transcurrido el experimento

% TODO: Añadir nuevo capítulo - Descripción de los resultados

% Aquí se detallan los resultados obtenidos, con tablas de resumen, gráficas de resultados, identificación de datos relevantes, etc.
% Es una exposición objetiva, sin valorar los resultados ni justificarlos.

@@

% Se aporta la discusión de los resultados presentados en el capítulo anterior. En este capítulo se puede discutir la relevancia de
% los resultados, presentar posibles explicaciones apra los datos anómalos y resaltar aquellos datos que sean parcitualarmente relevantes
% para el análisis del experimento.

@@

% Resumen final del trabajo y debe servir para informar del alcance y relevancia de la aportación.
% Suele estructurarse empezando con un resuemn del problema tratado, de cómo se ha abordado y de por qué la solución sería valída.

% Es recomendable que incluya también un resumen de las contribuciones del trabajo, en el que se relacionen las contribuciones y los
% resultado obtenidos con ls objetivos panteados para el trabajo, discutiendo hasta qué punto has conseguido resolver los objetivos planteados.

>>>>

\usepackage{apacite}
\usepackage{caption}
\usepackage{eso-pic}
\usepackage{float}
\usepackage{graphicx}
\usepackage{minted}
\usepackage{picture}
\usepackage{subcaption}
\usepackage[nottoc]{tocbibind}

@@

Stellar open clusters (OCs) \cite{janes1982open} are groups of stars gravitationally bound originated from a single molecular gas cloud.
Thus they share the same chemical composition and age, and they have similar relative positions and proper motion.
These astronomical objects are of fundamental importance to understand the spiral structure,
the dynamics and the chemical evolution of our galaxy.

@@

Young OCs allow us to research stars formation regions and improve our understanding about the mechanisms that create those stars.
On the other side, older OCs give us information about stellar processes and how the galactic disk evolves.
Some highly disturbed orbits could also provide evidence of recent merge events and accretion traces from outside the galaxy
\cite{cantat2016abundances}.

The study of OCs has been pushed forward thanks to the huge and precise dataset from the Gaia mission
\cite{collaboration2016description} Gaia DR2 \cite{gaia2018gaia}, available since 2018.
This dataset has helped to review already known open clusters and to find new ones.

Stars that belong to a same OC share relative positions, so their parallax value is similar among them. They also share similar values of
proper motion, both in right ascension and declination. Another property is that their metallicity must be uniform, since these stars were
born from the same gas cloud and at the same temporary stage. However, to take into account this last property, we face the inconvenient that this
parameter is not directly available in Gaia DR2 database. We will avoid this issue by looking at $E(b-r) / G_{mag}$ diagrams with well
defined isochrone curves \cite{bressan2012parsec}. These isochrone curves are derived from theoretical models that mainly involve metallicity
and mass/brightness from stars. With these models we can define important properties for the development of stars and their evolution in the
Hertzsprung–Russell (H-R) diagram. Examples of these curves are shown in Figure \ref{fig:examples_of_isochrones}.

@@

We will take advantage of these properties to validate our OCs characterization.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.9\textwidth]{../figures/theoretical_isochrones_in_hr_diagrams.pdf}
  \caption{Examples of theoretical isochrones in the H-R diagram taken from \protect\citeA[p.~16]{bressan2012parsec}}
  \label{fig:examples_of_isochrones}
\end{figure}

@@

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/melotte_22/pm_melotte_22.pdf}
      \caption{Proper motions}
      \label{fig:pm_melotte_22}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/melotte_22/parallax_melotte_22.pdf}
      \caption{Parallax}
      \label{fig:pm_vec_melotte_22}
    \end{subfigure}
  \end{subfigure}
  \caption{Open Cluster Melotte 22 (Messier 45)}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/ngc_6494/pm_ngc_6494.pdf}
      \caption{Proper motions}
      \label{fig:pm_ngc_6494}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/ngc_6494/parallax_ngc_6494.pdf}
      \caption{Parallax}
      \label{fig:parallax_ngc_6494}
    \end{subfigure}
  \end{subfigure}
  \caption{Open Cluster NGC 6494}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.5\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/melotte_22/raw_pm_melotte_22.pdf}
  \end{subfigure}
  \caption{Melotte 22 proper motions}
  \label{fig:raw_pm_melotte_22}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/melotte_22/raw_parallax_melotte_22.pdf}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/melotte_22/raw_hr_diagram_melotte_22.pdf}
    \end{subfigure}
  \end{subfigure}
  \caption{Melotte 22 parallax and H-R diagram}
  \label{fig:raw_parallax_hr_diagram_mellote_22}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/melotte_22/features_melotte_22.pdf}
  \end{subfigure}
  \caption{Melotte 22 feature pairplot}
  \label{fig:features_melotte_22}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans/kmeans_n2_pm_melotte_22.pdf}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans/kmeans_n5_pm_melotte_22.pdf}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans/kmeans_n8_pm_melotte_22.pdf}
    \end{subfigure}
  \end{subfigure}
  \medskip
  \begin{subfigure}{0.9\textwidth}
    \centering
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans/kmeans_n2_parallax_melotte_22.pdf}
      \caption{N clusters = 2}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans/kmeans_n5_parallax_melotte_22.pdf}
      \caption{N clusters = 5}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans/kmeans_n8_parallax_melotte_22.pdf}
      \caption{N clusters = 8}
    \end{subfigure}
  \end{subfigure}
  \caption{K-Means comparisons with Melotte 22}
  \label{fig:kmeans_comparisons_melotte_22}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \begin{subfigure}[t]{.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/melotte_22/kmeans_pm_melotte_22.pdf}
      \caption{Proper Motion}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/melotte_22/kmeans_parallax_melotte_22.pdf}
      \caption{Parallax}
    \end{subfigure}
  \end{subfigure}
  \caption{K-Means model applied to Melotte 22}
  \label{fig:kmeans_melotte_22}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.5\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/melotte_22/kmeans_hr_diagram_melotte_22.pdf}
  \end{subfigure}
  \caption{Melotte 22 H-R diagram with K-Means characterization}
  \label{fig:kmeans_hr_diagram_melotte_22}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/melotte_22/dec_pm_melotte_22.pdf}
      \caption{Proper Motion}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/melotte_22/dec_parallax_melotte_22.pdf}
      \caption{Parallax}
    \end{subfigure}
  \end{subfigure}
  \caption{DEC model applied to Melotte 22}
  \label{fig:dec_melotte_22}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.5\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/melotte_22/dec_hr_diagram_melotte_22.pdf}
  \end{subfigure}
  \caption{Melotte 22 H-R diagram with DEC characterization}
  \label{fig:dec_hr_diagram_melotte_22}
\end{figure}

@@

We have seen in the previous chapter that the DEC model is able to refine the clustering made by K-Means. This allows us to get
a more accurate characterization of the cluster.

In order to validate our model and be able to say how good or bad it is we need a different method to characterize the cluster
and compare one against each other.

For that reason we have used tools from the Virtual Observatory (VO) such as Clusterix and TOPCAT to characterize the studied clusters.

\section{Cluster Characterization with VO Tools}

\section{Comparison}

Until now, we have focus our study in Melotte 22 OC. Since our main aim is to get a model that fits well for a wide range of clusters,
we present now the results obtained for a selection of clusters with different topologies.

% TODO: Exposición de resultados y comentarios
% Es una exposición objetiva, sin valorar los resultados ni justificarlos.

@@

% Se aporta la discusión de los resultados presentados en el capítulo anterior. En este capítulo se puede discutir la relevancia de
% los resultados, presentar posibles explicaciones para los datos anómalos y resaltar aquellos datos que sean particularmente relevantes
% para el análisis del experimento.

@@

% Resumen final del trabajo y debe servir para informar del alcance y relevancia de la aportación.
% Suele estructurarse empezando con un resumen del problema tratado, de cómo se ha abordado y de por qué la solución sería válida.

% Es recomendable que incluya también un resumen de las contribuciones del trabajo, en el que se relacionen las contribuciones y los
% resultado obtenidos con ls objetivos planteados para el trabajo, discutiendo hasta qué punto has conseguido resolver los objetivos planteados.
